---
title: "Lab 4 Comments: Causation in Observational Studies"
subtitle: "Replicating Ferwerda and Miller (2014)"
author: "Jason Barajas"
date: "Last Updated `r format(Sys.Date())`"
output:
  html_document:
    theme: journal
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE
---



# Overview {-}

In this assignment, we'll walk through the logic and design of [Ferwerda and Miller (2014).](https://pols2580.paultesta.org/files/readings/ferwerda_miller_2014.pdf) 

## Goals

Conceptually, our goal in this lab is to see how scholars might use historical knowledge to make causal claims with observational data.

Specifically, we will see how F&M leverage a claim about how borders are drawn to assess the effects of different types of governing strategies.

Practically, we will continue to develop our statistical skills, introducing some core concepts from base `R`. 

Spefically we will see how we can use:

- `for()` loops to repeat a process like calculating a mean, over multiple variables
- self-defined functions to abstract and generalize repeated tasks
- the `with()` function to avoid having to write out `df$variable` 
- different types of `apply()` functions (namely `sapply()` and `tapply()`) to apply functions to a sets of variables (`sapply()`) and to subgroups within a set of variables (`tapply()`)

These are useful skills that broadly help you write your code more efficiently.  Things like `for()` loops, `functions()` and `apply()` can reduce the amount of copying, pasting and replacing you have to do, which in turn can reduce the amount of errors induced by forgetting to change a variable name, or mistyping a command.

But the first time you see a for loop, or define your own function, it will likely seem a bit abstract, and obtuse.That's ok. The goal is that you have a better, if not perfect, understanding of these concepts which we will use throughout the course. 

Further Notes 
- the with() function does not alter the original data frame at any cost. It gives the output separately for every value associated with the columns of the data frame.

# Read Ferwerda and Miller (2014) and answer the following:

- **What's the research question?** 
Do foreign occupiers face less resistance when they increase the level of native governing authority?

- **What's the theoretical framework?** 
The authors' theoretical framework holds that domestic groups seek political power in the present and respond to immediate threats to political survival; mass behavior under occupation is thus driven by short-term incentives rather than longterm strategic calculations concerning the international balance of power.

- **What's the empirical design?** 
The authors leverage a regression discontinuity design using a natural experiment based on the locally arbitrary assignment of French municipalities into German or Vichy-governed zones during World War II. They examine communes in close proximity to the demarcation line and measure differences in levels of resistance activity between the two zones.

- **What's are the main results?** 
Domestic groups that were granted governing authority were less likely to engage in resistance activity, while violent resistance was heightened in regions dominated by groups excluded from the governing regime. 


# Load the replication data and set up your workspace

Please run the following command to load the data:


```{r}
# Load the data
load(url("https://pols2580.paultesta.org/files/data/04_lab.rda"))
```

And write code to load any packages you think you might need to analyze the data.

```{r libraries}
# Libraries
the_packages <- c(
  ## R Markdown
  "kableExtra","DT",
  ## Tidyverse
  "tidyverse", "lubridate", "forcats", 
  "haven", "labelled",
  ## Extensions for ggplot
  "ggmap","ggrepel", "ggridges", 
  "ggthemes", "ggpubr", "GGally",
  "scales", "dagitty", "ggdag")
```


The next commands tell us a little about the dataset. Specifically, there are 1371 observations (rows) and 14 variables (columns)

```{r}
dim(wrk)
names(wrk)
```

Here's a brief codebook to help you along the way


- **south**: 0-1 indicator, 1 if south of demarcation line (Vichy), 0 otherwise (Germany)
- **distancetotrain**: Distance to nearest train station from center of commune
- **communications**: Ordinal variable treated as interval indicating whether a commune had a telephone bureau (3?), a telegraph station (2?), or a post office (1?) or none of these (0?) in 1939
- **hectares_scaled**: number of actively farmed hectares as a proxy for forest cover and urbanization
- **population**: Population from the 1936 census
- **pctright**: Vote shares for the Right from the 1936 national legislative elections 
- **pctleft**: Vote shares for the Left from the 1936 national legislative elections
- **sd_elev250**: The standard deviation of elevation using resampled 250m resolution SRTM elevation data; a proxy for Ruggedness of Terrain
- **near_km**: Distance, in kilometers to the Demarcation line
- **s**: Count of the number of sabotage activities (all attacks against infrastructure (largely railroad and communications)
- **fighting**:: Count of the number of resistance-initiated attacks on German and Vichy personnel
- **area_sqkm**: Area in square kilometers
- **north**: 0-1 indicator, 1 if north of demarcation line (Germany), 0 otherwise (Vichy)
- **nom\_dept**: Name of the department.


-------------------------

# What's the unit of analysis for this study?

**Answer:** 
The unit of analysis are geographical units-- the north and south sides of the demarcation line in the selected 4 departments. 

# Describe the structure of the data

Specifically:

- How many departments are there in the sample? 
```{r}
table(wrk$nom_dept)
```
*Answer*
There are four (4) departments in the sample. 

- How many communes in each department are north and south of the demarcation line?

```{r}
south_communes <- table(wrk$nom_dept, wrk$south == 1)
south_communes
```
*Answer*
62 communes south of the demarcation line in Charente
299 communes north of the demarcation line in Charente 

116 communes south of the demarcation line in Cher
133 communes north of the demarcation line in Cher 

336 communes south of the demarcation line in Saone-et-Loire 
175 communes north of the demarcation line in Saone-et-Loire 

57 communes south of the demarcation line in Vienne
193 communes north of the demarcation line in Vienne 


```{r}
# Write your code here
dim(wrk)
```

**Answer:** 
There are four departments in the study. There are 1371 communes.


# Describe the study's outcomes of interest and their distributions 


```{r}
# Write your code here
summary(wrk)
```

**Answer:** 
The study's outcomes of interest and their distributions include: 
"pctright", which ranges from 0 to 0.8246 with a median of 0.3643 and a mean of 0.3459. 
"pctleft", which ranges from 0.1021 and 0.9196 with a median of 0.4814 and a mean of 0.4515.
"s", which ranges from 0 to 29 with a median of 0 and a mean of 0.4362. 
"fighting", which ranges from 0 to 19 with a median of 0 and a mean of 0.434. 

-------------------------

# What is the "treatment" in this study? 



```{r}
# Write your code here
table(wrk$north)
table(wrk$south)
```

**Answer:** 
What variable or variables in the data describe treatment status?
"south" indicates treatment status. Treatment is living in a Vichy-controlled commune.    


-------------------------

So one of the implications of this design is that a commune's status relative to the demarcation line (and thus type of government/devolution of power) is "as-if" random. There's no way to really prove this claim, but we can test some of its observable implications. 

Specifically, if the Germans essentially drew the demarcation line at random (at least relative to the local boundaries of specific communes), then the observable characteristics of communes north of the line (under German occupation) and south of the line (under Vichy rule) should be similar. 


# Calculate the means and differences of means for all communes north and south of the demarcation line for the following variables:

- **distancetotrain**: Distance to nearest train station from center of commune
- **communications**: Ordinal variable treated as interval indicating whether a commune had a telephone bureau (3?), a telegraph station (2?), or a post office (1?) or none of these (0?) in 1939
- **hectares_scaled**: number of actively farmed hectares as a proxy for forest cover and urbanization
- **population**: Population from the 1936 census
- **pctright**: Vote shares for the Right from the 1936 national legislative elections 
- **pctleft**: Vote shares for the Left from the 1936 national legislative elections
- **sd_elev250**: The standard deviation of elevation using resampled 250m resolution SRTM elevation data; a proxy for Ruggedness of Terrain
- **near_km**


```{r}
# Write your code here
# You might find something like the following useful

# # List of covariates
the_vars<-c("distancetotrain","communications","hectares_scaled","population",
"pctright","pctleft","sd_elev250","near_km")

# # Holders for means
north_means<-c()
south_means<-c()

# #Replace ???? in code below with some useful logical statement

for(i in the_vars){ #applies to all individual observations in our function             
     print(paste("Now summarizing",i))
     north_means[i]<-mean(wrk[wrk$north == 1,i])
     south_means[i]<-mean(wrk[wrk$south == 1,i])

}
south_means
north_means

#Finding the difference between the means 
dif_means <-north_means - south_means
dif_means
```


**Answer:**
Means for Covariates South of Demarcation Line: 
distancetotrain  communications hectares_scaled      population        pctright         pctleft      sd_elev250 
      4.2889667       1.4956217      39.6845843     740.1295972       0.2743438       0.5333780      25.5892396 
        near_km 
     19.8388128 

Means for Covariates North of the Demarcation Line 
distancetotrain  communications hectares_scaled      population        pctright         pctleft      sd_elev250 
      4.0037500       1.5375000      51.2661532     997.4450000       0.3969024       0.3930110      20.3176547 
        near_km 
     23.9157269 

Differences in Means: 
distancetotrain  communications hectares_scaled      population        pctright         pctleft      sd_elev250 
    -0.28521673      0.04187828     11.58156888    257.31540280      0.12255864     -0.14036706     -5.27158485 
        near_km 
     4.07691405 



-------------------------

So how should we make sense of these differences? Are they substantively large? Do they provide support for or against a claim that the demarcation lines is "as if" random?

Answering these questions has proved difficult, since different variables are measured on different scales, and it's hard to quantify what big is. 

Last week, we saw that we could assess these differences relative to a claim about the world ("i.e. if there were no difference between theses groups, how likely is it that we would a difference as large as we did").

Today, we'll consider another way of thinking about how big differences are, which is to think about how much the difference might vary. 

We know from our first week on measurement, that things like variance and standard deviations can be used to describe the spread of a random variable's distribution (how much observations tend to vary around their means). 

Means and differences of means are functions of variables, and so we can think about these quantities as having their own distributions and spreads.

So one way to compare the magnitude of the differences you calcuated above, is to standardize each difference by its standard error -- that is, we could divide each difference by the standard deviation of the sampling distribution of that difference (how much we might expect that difference to vary). Don't worry about the terminology right now--we'll cover these concepts in greater detail later in the course. 

For now, we'll use as a rough rule of thumb, that **standardized differences greater than 2 are "big,"** that is, they are unlikely to arise by chance and thus provide evidence against the claim that the demarcation line was drawn as if random.

The code below **defines a function** to calculate these standardized differences and returns the standardized difference (as well as the means, difference of means, and standard error of that difference) for the `distancetotrain` variable.

```{r}

std_diff_fn <- function(x){
  # inputs
  # x: the variable we want to calculate a standardized difference of
  stopifnot(is.character(x))
  
  #1.  Calculate means and difference of means
  mu1 <- with(wrk, mean(wrk[north==1, x],na.rm=T)) #means when the north equals 1
  mu0 <- with(wrk, mean(wrk[north==0, x],na.rm=T)) #means when the north equals 0
  diff <- mu1 - mu0 #the difference in means for the north

  #2. Calculate "standard error" of difference
      # Assuming unequal variance of groups
  sd1 <- with(wrk, sd(wrk[north==1, x],na.rm=T))
  sd0 <- with(wrk, sd(wrk[north==0, x],na.rm=T))
  n1 <- with(wrk, sum(!is.na(wrk[north==1, x])))
  n0 <- with(wrk, sum(!is.na(wrk[north==0, x])))
  se <- sqrt( sd1^2/n1 + sd0^2/n0)

  #3. Standardize difference
  std_diff <- diff/se
  
  #4. Combine Results
  results <- c(North=mu1, South=mu0,Differnece=diff, SE = se, StdDiff = std_diff)
  return(results)
  
}

std_diff_fn("distancetotrain")


```


# Use the `sapply()` function to apply, the `std_diff_fn` to the covariates in the object `the_vars.` 

Evaluate these standardized differences in relation to the authors' claims that the demarcation line can be seen as being as-if randomly assigned.


```{r}
# Write your code here
# hint:
# sapply(some object here, some function here)

sapply(the_vars, std_diff_fn)
```

**Answer:**
Hectares_scaled, population, pctleft, sd_elev250 and near_km all have standard differences higher than 2. This means that there are insignificant differences between the two groups in more than half of the selected covariates, signifying that they may be good counterfactuals of each other. 


-----------------------------

Hmmm... so using our "standardized differences bigger than two" rule of thumb, we became a little bit skeptical about whether the claim of as-if random was justified for the whole sample.

Now we'll explore whether for a subset of the sample, this claims appears more plausible. 

Specifically, we'll look at whether for communes within a narrower window or bandwidth around the demarcation line differences in observed covariates between communes north and south of the line are small, and thus seem consistent with a claim that the line of demarcation is as good as randomly assigned for these communes.

To do so, we'll modify our `std_diff_fn` and create a new function called `new_std_diff_fn`


```{r}
new_std_diff_fn <- function(x, c = max(wrk$near_km,na.rm=T)){
  # inputs
  # x: the variable we want to calculate a standardized difference of
  # c: the cutoff distance from the border
  stopifnot(is.character(x))
  stopifnot(is.numeric(c))

  # What's going on here?
  wrk <- wrk[wrk$near_km <= c,]
  
  #1.  Calculate means and difference of means
  mu1 <- with(wrk, mean(wrk[north==1, x],na.rm=T))
  mu0 <- with(wrk, mean(wrk[north==0, x],na.rm=T))
  diff <- mu1 - mu0

  #2. Calculate "standard error" of difference
      # Assuming unequal variance of groups
  sd1 <- with(wrk, sd(wrk[north==1, x],na.rm=T))
  sd0 <- with(wrk, sd(wrk[north==0, x],na.rm=T))
  n1 <- with(wrk, sum(!is.na(wrk[north==1, x])))
  n0 <- with(wrk, sum(!is.na(wrk[north==0, x])))
  se <- sqrt( sd1^2/n1 + sd0^2/n0)

  #3. Standardize difference
  std_diff <- diff/se
  
  #4. Combine Results
  results <- c(North=mu1, South=mu0,Differnece=diff, SE = se, StdDiff = std_diff)
  return(results)
  
}

new_std_diff_fn("distancetotrain", c=10)

```


# Explain how `new_std_diff_fn()` differences from `std_diff_fn()`



**Answer:** 
The new function 'new_std_diff_fn()' is measuring the standard differences across the north and south groups but at a narrower scope-- 10 kilomteres within the demarcation line. This will hopefully produce a sample of observations that are more similar to each other and be good counterfactuals for each other. 

-------------------------

Great! So this modified function can calculate conditional means for a subset of observations (communes) within `c` kilometers of the demarcation line.


# Does this claim appear more or less plausible within 10 km of the line?

Use  `sapply()` with `new_std_diff_fn()` to calculate the standardized differences for the covariates in `the_vars` for communes within 10 km north or south of the demarcation line (i.e., c=10). 

Evaluate these differences as they relate to F&M's claim of "as-if" randomization. 


```{r}
# Write your code here

sapply(the_vars, new_std_diff_fn)
```



**Answer:** 
Looking at observations within 10km we see that the assumptions seem more plausible, with the exception of the  communications and distance to train covariates. 

-------------------------

So the identifying assumption seems more plausible for communes closer to the demarcation line. Now let's estimate the "effect" of German rule on resistance activity (sabotage and fighting) for the full sample and for the subset of communes within 10 km of the demarcation line.


# Does devolution reduce resistance?


Calculate the difference in the mean number of sabotage events and fighting events for communes north and south of the demarcation line, for the full sample of communes and the communes within 10 km of the demarcation line. Provide a substantive interpretation of these differences.

You can use logical indexing, or the `new_std_diff_fn` if you'd like. Try to do it both ways to savor the joy of logical indexing and the magic of functions.

```{r}
# Write your code here
mu_north_s <- with(wrk,mean(wrk$s[north==1 & wrk$near_km <= 10]))

mu_south_s <- with(wrk,mean(wrk$s[south==1 & wrk$near_km <= 10]))

diff_mean_sab <- mu_north_s - mu_south_s 

diff_mean_sab
```

**Answer:**
The standard difference in means for the sabotage covariate leads us to believe that there is no statistical significance to the  results because it is less than 2. 



-------------------------


# Repeat the calculations above, but this time comparing the differences between communes north and south of the line within the same department. 

Comment on the relative similarity of or differences in the findings from estimating effects for each department compared to pooling all the communes together?

Below is some code to get you started the hard way, you could also try modifying the `new_std_diff_fn`

```{r}
# Here's some code to get you started

# 10k
# Sabotage
# Calculate means by department for communes in the north within 10 km of line
mus_s_north_dep_10k <- with(wrk[wrk$north==1 & wrk$near_km<=10,], tapply(s, nom_dept, mean)) 

# Calculate means by department for communes in the south within 10 km of line
mus_s_south_dep_10k <- with(wrk[wrk$north==0 & wrk$near_km<=10,], tapply(s, nom_dept, mean))

# Calculate differences within each department
diffs_s_south_dep_10k <- mus_s_north_dep_10k-mus_s_south_dep_10k

# Print differences.
diffs_s_south_dep_10k

###########
# Your turn
###########

# 10k
# Fighting
# Calculate means by department for communes in the north within 10 km of line
mus_fighting_north_dep_10k <- with(wrk[wrk$north==1 & wrk$near_km<=10,], tapply(fighting, nom_dept, mean)) 
# Calculate means by department for communes in the south within 10 km of line
mus_fighting_south_dep_10k <- with(wrk[wrk$north==0 & wrk$near_km<=10,], tapply(fighting, nom_dept, mean))

diffs_fighting_south_dep_10k <- mus_fighting_north_dep_10k - mus_fighting_south_dep_10k

#Print the differences 
diffs_fighting_south_dep_10k

# Full sample 
# Sabotage
# Calculate means by department for communes in the north 
mus_s_north_dep_full <- with(wrk[wrk$north==1,], tapply(s, nom_dept, mean)) 

# Calculate means by department for communes in the south 
mus_s_south_dep_full <- with(wrk[wrk$north==0,], tapply(s, nom_dept, mean))

diffs_fighting_south_dep_full <- mus_s_north_dep_full - mus_s_south_dep_full

#Print the differences 
diffs_fighting_south_dep_full


# Full sample 
# Fighting
mus_fighting_north_dep_full <- with(wrk[wrk$north==1,], tapply(fighting, nom_dept, mean)) 
# Calculate means by department for communes in the south within 10 km of line
mus_fighting_south_dep_full <- with(wrk[wrk$north==0,], tapply(fighting, nom_dept, mean))

diffs_fighting_south_dep_full <- mus_fighting_north_dep_full - mus_fighting_south_dep_full

#Print the differences 
diffs_fighting_south_dep_full
```



**Answer:** 
There are some minor changes in statistical significance compared to differences observed at the commune level. However, they are not statistically significant as they are not above 2. 
